---
title: "PamChip Kinome Array Annotation Report"
subtitle: "Family-Level KRSA Mapping, Coverage, and Multi-Source Evidence Summary"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    embed-resources: true
    grid:
      sidebar-width: 250px
      body-width: 900px
---

```{r setup, include=FALSE}
library(knitr)
library(targets)
library(dplyr)
library(ggplot2)
library(kableExtra)
library(tidyr)
library(stringr)

opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 9,
  fig.height = 5,
  fig.align = "center"
)

store_path <- here::here("_targets")

manifest                       <- tar_read(manifest, store = store_path)
validated_chip_sites           <- tar_read(validated_chip_sites, store = store_path)
master_evidence                <- tar_read(master_evidence, store = store_path)
family_evidence                <- tar_read(family_evidence, store = store_path)
family_mapping                 <- tar_read(family_mapping, store = store_path)
experimental_mapping           <- tar_read(experimental_mapping, store = store_path)
kinase_library_mapping         <- tar_read(kinase_library_mapping, store = store_path)
gps6_mapping                   <- tar_read(gps6_mapping, store = store_path)
predictive_concordant_mapping  <- tar_read(predictive_concordant_mapping, store = store_path)
concordant_mapping             <- tar_read(concordant_mapping, store = store_path)
qc_summary                     <- tar_read(qc_summary, store = store_path)
coverage_summary               <- tar_read(coverage_summary, store = store_path)

value_or_na <- function(x, name) {
  if (is.list(x) && name %in% names(x)) x[[name]] else NA
}

pct <- function(x) {
  ifelse(
    is.na(x),
    NA_character_,
    paste0(round(100 * x, 1), "%")
  )
}

first_existing <- function(data, candidates) {
  hit <- candidates[candidates %in% names(data)]
  if (length(hit) == 0L) return(NULL)
  hit[[1L]]
}
```

# Executive Summary

This report summarizes the reproducible kinase-substrate annotation workflow for the configured PamChip layout. The pipeline integrates three complementary evidence streams:

1. **PhosphoSIGNOR** — experimentally curated kinase-substrate relationships.
2. **GPS 6.0** — sequence-based predictive evidence using the local GPS6 model hierarchy.
3. **The Kinase Library** — kinase motif percentile predictions.

Source-native kinase evidence is preserved in the master evidence table, while the primary KRSA mapping is collapsed to the **kinase-family level** to reflect the practical resolution of the PamChip assay.

```{r executive-summary}
chip_name <- manifest$chip$name %||% manifest$chip$type %||% basename(manifest$chip$file)

summary_df <- tibble::tibble(
  Metric = c(
    "Chip / layout",
    "Validated chip rows",
    "Master evidence rows",
    "Family evidence rows",
    "Primary KRSA mapping rows",
    "Experimental-only mapping rows",
    "Kinase Library-only mapping rows",
    "GPS6-only mapping rows",
    "Predictive-concordant mapping rows",
    "Multi-source concordant mapping rows"
  ),
  Value = c(
    chip_name,
    nrow(validated_chip_sites),
    nrow(master_evidence),
    nrow(family_evidence),
    nrow(family_mapping),
    nrow(experimental_mapping),
    nrow(kinase_library_mapping),
    nrow(gps6_mapping),
    nrow(predictive_concordant_mapping),
    nrow(concordant_mapping)
  )
)

kable(summary_df, align = c("l", "r")) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  )
```

# Array Layout and Coverage

```{r coverage-summary}
coverage_table <- NULL

if (is.data.frame(coverage_summary)) {
  coverage_table <- coverage_summary
} else if (is.list(coverage_summary)) {
  scalar_values <- vapply(
    coverage_summary,
    function(x) length(x) == 1L && (is.atomic(x) || is.character(x)),
    logical(1)
  )

  if (any(scalar_values)) {
    coverage_table <- tibble::tibble(
      Metric = names(coverage_summary)[scalar_values],
      Value = vapply(
        coverage_summary[scalar_values],
        function(x) as.character(x[[1L]]),
        character(1)
      )
    )
  }
}

if (!is.null(coverage_table) && nrow(coverage_table) > 0L) {
  kable(coverage_table) %>%
    kable_styling(
      bootstrap_options = c("striped", "hover", "condensed"),
      full_width = FALSE
    )
} else {
  cat("Coverage metrics were generated by the pipeline but are not stored as a directly tabular object.\n")
}
```

## Peptide Mapping Coverage

```{r peptide-coverage}
peptide_col <- first_existing(
  validated_chip_sites,
  c("peptide_id", "PeptideID", "ID", "id")
)

mapping_peptide_col <- first_existing(
  family_mapping,
  c("peptide_id", "PeptideID", "ID", "id")
)

if (!is.null(peptide_col) && !is.null(mapping_peptide_col)) {
  total_peptides <- validated_chip_sites %>%
    distinct(.data[[peptide_col]]) %>%
    nrow()

  mapped_peptides <- family_mapping %>%
    distinct(.data[[mapping_peptide_col]]) %>%
    nrow()

  coverage_pct <- if (total_peptides > 0L) {
    mapped_peptides / total_peptides
  } else {
    NA_real_
  }

  coverage_df <- tibble::tibble(
    Metric = c(
      "Unique peptides on chip",
      "Unique peptides in primary mapping",
      "Primary mapping coverage"
    ),
    Value = c(
      total_peptides,
      mapped_peptides,
      pct(coverage_pct)
    )
  )

  kable(coverage_df, align = c("l", "r")) %>%
    kable_styling(
      bootstrap_options = c("striped", "hover"),
      full_width = FALSE
    )
}
```

# Evidence Composition

The master evidence table preserves the source-native evidence before family-level collapse.

```{r evidence-source-counts}
if ("evidence_source" %in% names(master_evidence)) {
  source_counts <- master_evidence %>%
    count(evidence_source, name = "Evidence_Rows", sort = TRUE)

  kable(
    source_counts,
    col.names = c("Evidence Source", "Evidence Rows"),
    align = c("l", "r")
  ) %>%
    kable_styling(
      bootstrap_options = c("striped", "hover"),
      full_width = FALSE
    )
}
```

```{r evidence-source-plot}
if ("evidence_source" %in% names(master_evidence)) {
  master_evidence %>%
    count(evidence_source, name = "Evidence_Rows") %>%
    ggplot(
      aes(
        x = reorder(evidence_source, Evidence_Rows),
        y = Evidence_Rows
      )
    ) +
    geom_col() +
    coord_flip() +
    theme_minimal(base_size = 13) +
    labs(
      title = "Evidence Rows by Source",
      x = "Evidence Source",
      y = "Number of Evidence Rows"
    )
}
```

# Family-Level Evidence

The following table summarizes the kinase families represented after source-native evidence is collapsed to the resolution used for KRSA mapping.

```{r top-families}
family_col <- first_existing(
  family_evidence,
  c("kinase_family", "family", "KinaseFamily", "kinase")
)

family_peptide_col <- first_existing(
  family_evidence,
  c("peptide_id", "PeptideID", "ID", "id")
)

if (!is.null(family_col)) {
  top_families <- family_evidence %>%
    filter(!is.na(.data[[family_col]])) %>%
    group_by(.data[[family_col]]) %>%
    summarise(
      Evidence_Rows = n(),
      Target_Peptides = if (!is.null(family_peptide_col)) {
        n_distinct(.data[[family_peptide_col]])
      } else {
        NA_integer_
      },
      .groups = "drop"
    ) %>%
    arrange(desc(Target_Peptides), desc(Evidence_Rows)) %>%
    slice_head(n = 20)

  names(top_families)[1] <- "Kinase_Family"

  kable(
    top_families,
    col.names = c(
      "Kinase Family",
      "Evidence Rows",
      "Target Peptides"
    ),
    align = c("l", "r", "r")
  ) %>%
    kable_styling(
      bootstrap_options = c("striped", "hover", "condensed"),
      full_width = FALSE
    )
}
```

```{r top-families-plot}
if (exists("top_families") && nrow(top_families) > 0L) {
  ggplot(
    top_families,
    aes(
      x = reorder(Kinase_Family, Target_Peptides),
      y = Target_Peptides
    )
  ) +
    geom_col() +
    coord_flip() +
    theme_minimal(base_size = 12) +
    labs(
      title = "Top Kinase Families by Mapped Chip Peptides",
      x = "Kinase Family",
      y = "Unique Target Peptides"
    )
}
```

# Source-Specific Mapping Sensitivity

The primary mapping is accompanied by sensitivity mappings that isolate the contribution of each evidence source and the degree of agreement between sources.

```{r mapping-sensitivity}
mapping_sensitivity <- tibble::tibble(
  Mapping = c(
    "Primary family mapping",
    "Experimental only (PhosphoSIGNOR)",
    "Kinase Library only",
    "GPS6 only",
    "Predictive concordance (Kinase Library + GPS6)",
    "Multi-source concordance"
  ),
  Rows = c(
    nrow(family_mapping),
    nrow(experimental_mapping),
    nrow(kinase_library_mapping),
    nrow(gps6_mapping),
    nrow(predictive_concordant_mapping),
    nrow(concordant_mapping)
  )
)

kable(mapping_sensitivity, align = c("l", "r")) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width = FALSE
  )
```

```{r mapping-sensitivity-plot}
ggplot(
  mapping_sensitivity,
  aes(
    x = reorder(Mapping, Rows),
    y = Rows
  )
) +
  geom_col() +
  coord_flip() +
  theme_minimal(base_size = 12) +
  labs(
    title = "KRSA Mapping Size Across Evidence Rules",
    x = NULL,
    y = "Mapped Family-Peptide Rows"
  )
```

# GPS 6.0 Evidence

GPS6 output is retained at its native hierarchical resolution before downstream family collapse. This preserves broad family/subfamily models as well as deeper kinase-specific nodes where available.

```{r gps6-summary}
gps6_rows <- master_evidence %>%
  filter(evidence_source == "GPS6")

if (nrow(gps6_rows) > 0L) {
  gps6_summary <- gps6_rows %>%
    summarise(
      Evidence_Rows = n(),
      Unique_Nodes = n_distinct(gps6_node, na.rm = TRUE),
      Supported_Rows = sum(gps6_supported %in% TRUE, na.rm = TRUE),
      Median_Score = median(gps6_score, na.rm = TRUE)
    )

  kable(gps6_summary, digits = 4) %>%
    kable_styling(
      bootstrap_options = c("striped", "hover"),
      full_width = FALSE
    )
}
```

## Most Frequently Assigned GPS6 Nodes

```{r gps6-top-nodes}
if (exists("gps6_rows") && nrow(gps6_rows) > 0L && "gps6_node" %in% names(gps6_rows)) {
  gps6_top_nodes <- gps6_rows %>%
    filter(!is.na(gps6_node)) %>%
    count(gps6_node, sort = TRUE, name = "Evidence_Rows") %>%
    slice_head(n = 20)

  kable(gps6_top_nodes, align = c("l", "r")) %>%
    kable_styling(
      bootstrap_options = c("striped", "hover"),
      full_width = FALSE
    )
}
```

# Primary KRSA Mapping Preview

```{r family-mapping-preview}
preview_mapping <- family_mapping %>%
  slice_head(n = 30)

kable(preview_mapping) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    font_size = 11
  ) %>%
  scroll_box(
    width = "100%",
    height = "500px"
  )
```

# Master Evidence Preview

```{r master-evidence-preview}
preview_evidence <- master_evidence %>%
  slice_head(n = 30)

kable(preview_evidence) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    font_size = 10
  ) %>%
  scroll_box(
    width = "100%",
    height = "500px"
  )
```

# Quality Control

```{r qc-summary}
qc_table <- NULL

if (is.data.frame(qc_summary)) {
  qc_table <- qc_summary
} else if (is.list(qc_summary)) {
  scalar_values <- vapply(
    qc_summary,
    function(x) length(x) == 1L && (is.atomic(x) || is.character(x)),
    logical(1)
  )

  if (any(scalar_values)) {
    qc_table <- tibble::tibble(
      Metric = names(qc_summary)[scalar_values],
      Value = vapply(
        qc_summary[scalar_values],
        function(x) as.character(x[[1L]]),
        character(1)
      )
    )
  }
}

if (!is.null(qc_table) && nrow(qc_table) > 0L) {
  kable(qc_table) %>%
    kable_styling(
      bootstrap_options = c("striped", "hover", "condensed"),
      full_width = FALSE
    )
} else {
  cat("QC results were generated by the pipeline but are not stored as a directly tabular object.\n")
}
```

# Pipeline Configuration

```{r pipeline-configuration}
config_df <- tibble::tibble(
  Parameter = c(
    "GPS6 threshold",
    "GPS6 model scope",
    "GPS6 model directory",
    "Kinase Library percentile cutoff",
    "PhosphoSIGNOR force update"
  ),
  Setting = c(
    manifest$sources$gps6$threshold %||% "h",
    manifest$sources$gps6$model_scope %||% "ALL",
    manifest$sources$gps6$pre_dir %||% "data/pre",
    manifest$sources$kinase_library$percentile_cutoff %||% 0.90,
    manifest$sources$signor$force_update %||% FALSE
  )
)

kable(
  config_df,
  col.names = c("Pipeline Parameter", "Configured Value"),
  align = c("l", "r")
) %>%
  kable_styling(
    bootstrap_options = c("bordered", "condensed"),
    full_width = FALSE
  )
```

# Reproducibility Notes

The report is generated directly from the cached `targets` objects produced by the same pipeline run. GPS6 model files and the local GPS6 runner are tracked as explicit file dependencies, so changes to the predictor implementation or model/resource files invalidate the appropriate downstream targets.

The master evidence table preserves source identity and native GPS6 hierarchy. Family-level collapse occurs only after evidence harmonization, allowing the final KRSA mapping to remain conservative while retaining full provenance for sensitivity analyses.
