---
title: "PamChip Kinome Array Annotation Report"
subtitle: "Target Coverage, Kinase-Substrate Mappings, and Multi-Source Evidence Summary"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    embed-resources: true
    grid:
      sidebar-width: 250px
      body-width: 900px
---

```{r setup, include=FALSE}
library(knitr)
library(targets)
library(dplyr)
library(ggplot2)
library(kableExtra)
library(tidyr)

opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 9,
  fig.height = 5,
  fig.align = "center"
)

# Resolve root _targets store path dynamically
store_path <- here::here("_targets")

# Load pipeline deliverables from targets cache
manifest        <- tar_read(manifest, store = store_path)
parsed_sites    <- tar_read(parsed_chip_sites, store = store_path)
final_map       <- tar_read(final_kinase_substrate_map, store = store_path)
qc_summary      <- tar_read(qc_summary, store = store_path)
```
## Executive Summary
This report provides a automated summary of the kinase-substrate annotation pipeline for r manifest$project$name (r manifest$chip$type Array layout). Target peptides extracted from the array layout are mapped across three complementary evidence streams:

1. PhosphoSIGNOR: Curated literature interactions.

2. GPS 6.0: Sequence PSSM score predictions.

3. The Kinase Library: PSSM percentiles computed via Python.

## Array Layout & Target Coverage
``` {r meta}
total_peptides  <- qc_summary$total_input_peptides
mapped_peptides <- qc_summary$mapped_peptides
coverage_pct    <- qc_summary$coverage_percentage

coverage_df <- data.frame(
  Metric = c("Chip Type", "Total Peptides Extracted", "Mapped Peptides", "Target Coverage Rate"),
  Value  = c(manifest$chip$type, total_peptides, mapped_peptides, paste0(coverage_pct, "%"))
)

kable(coverage_df, col.names = c("Pipeline Metric", "Value"), align = c("l", "r")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

::: {.callout-note}

Coverage Overview
A total of r mapped_peptides out of r total_peptides target phosphopeptides on the array (r coverage_pct%) were mapped to at least one upstream kinase with high confidence.
:::

## Evidence Multiplicity Breakdown

The plot and table below highlight the degree of evidence overlap across database sources (PhosphoSIGNOR, GPS 6.0, and The Kinase Library).

``` {r evidence-multiplicity}
evidence_summary <- final_map %>%
  group_by(evidence_count) %>%
  summarise(Pair_Count = n(), .groups = "drop") %>%
  mutate(Evidence_Category = paste0(evidence_count, " Source(s)"))

ggplot(evidence_summary, aes(x = Evidence_Category, y = Pair_Count, fill = Evidence_Category)) +
  geom_bar(stat = "identity", width = 0.6, show.legend = FALSE) +
  scale_fill_brewer(palette = "Set2") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Kinase-Substrate Predictions by Evidence Multiplicity",
    x = "Evidence Tier",
    y = "Number of Mapped Interactions"
  )
```
``` {r evidence-multiplicity-kable}
overlap_df <- data.frame(
  Source = c("PhosphoSIGNOR (Curated)", "GPS 6.0 (Predicted)", "Kinase Library (Predicted)"),
  Supported_Pairs = c(
    qc_summary$source_overlaps$signor_supported,
    qc_summary$source_overlaps$gps6_supported,
    qc_summary$source_overlaps$kinase_library_supported
  )
)

kable(overlap_df, col.names = c("Evidence Stream", "Mapped Kinase-Substrate Pairs"), align = c("l", "r")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## Top Mapped Upstream Kinases

The top 15 upstream kinases ranked by total assigned array targets and mean ensemble scores:

``` {r top-kinases}
top_kinases <- final_map %>%
  group_by(kinase_clean) %>%
  summarise(
    Target_Peptides = n_distinct(id),
    Mean_Ensemble_Score = round(mean(ensemble_score, na.rm = TRUE), 3),
    .groups = "drop"
  ) %>%
  slice_max(order_by = Target_Peptides, n = 15)

ggplot(top_kinases, aes(x = reorder(kinase_clean, Target_Peptides), y = Target_Peptides, fill = Mean_Ensemble_Score)) +
  geom_col() +
  coord_flip() +
  scale_fill_viridis_c(option = "magma", name = "Mean Score") +
  theme_minimal(base_size = 12) +
  labs(
    title = "Top 15 Kinases by Mapped Substrate Targets",
    x = "Kinase Symbol",
    y = "Number of Mapped Array Peptides"
  )
```

## Sample Mappings Preview
A sample of top-scoring integrated kinase-substrate interactions:

``` {r sample-mappings}
sample_data <- final_map %>%
  select(
    `Peptide ID` = id,
    `UniProt` = substrate_uniprot,
    `Site` = res_position,
    `Kinase` = kinase_clean,
    `SIGNOR` = SIGNOR,
    `GPS 6.0` = `GPS6.0`,
    `Kinase Lib` = KinaseLibrary,
    `Ensemble Score` = ensemble_score
  ) %>%
  mutate(`Ensemble Score` = round(`Ensemble Score`, 3)) %>%
  head(20)

kable(sample_data, align = "c") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), font_size = 12) %>%
  scroll_box(width = "100%", height = "400px")
Pipeline Configuration Settings
Code snippet
config_df <- data.frame(
  Parameter = c("GPS 6.0 Cutoff Threshold", "Kinase Library Percentile Cutoff", "SIGNOR Weight", "GPS 6.0 Weight", "Kinase Library Weight"),
  Setting   = c(
    manifest$sources$gps6$threshold,
    manifest$sources$kinase_library$cutoff,
    manifest$mapping_weights$signor,
    manifest$mapping_weights$gps6,
    manifest$mapping_weights$kinase_library
  )
)

kable(config_df, col.names = c("Pipeline Parameter", "Configured Value"), align = c("l", "r")) %>%
  kable_styling(bootstrap_options = c("bordered", "condensed"), full_width = FALSE)
```
